---
title: "01_MDD_Analysis"
author: "Sonick Suri"
date: "2024-07-08"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 4
    toc_float: true
---

## 1. Introduction

This notebook contains entire Analysis to generate insights using data from routine clinical practice to better understand the treatment of major depressive disorder (MDD). MDD is typically treated with antidepressants and psychological therapy.

This notebooks Contains following items

* Exploration of the data

* Data Processing

* Cohort Creation

* Data Analysis

## 2. Data Exploration

### 2.1 Import Necessary Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(rtables)
library(scales)
library(broom)
```

### 2.2 Load dataset

```{r}
bill_amount <- read.csv("bill_amount.csv")
bill_id <- read.csv("bill_id.csv")
clinical_data <- read.csv("clinical_data.csv")
demographics <- read.csv("demographics.csv")
```

### 2.3 Explore dataset

```{r}
View(bill_amount)
View(bill_id)
View(clinical_data)
View(demographics)
```

#### 2.3.1 Check Structure of datasets

Check structure of dataset to get idea of variables in each file. Also check if there are any formatting changes we will need to make in future.

```{r}
datasets <- list(bill_amount,bill_id,clinical_data,demographics)
data_str <- lapply(datasets, str)
print(data_str)
```

#### 2.3.2 Check Missing values in all variables

```{r}
# Create function to check missing value % in each column
missing_percentage <- function (data){
  round(colSums(is.na(data)) / nrow(data) * 100,2)
}

lapply(datasets, missing_percentage)
```


#### 2.3.3 Check Univariate distributions of all variables

This is to ensure all variables have been correctly populated and check if there are any outliers/cleaning required

Explore Histogram of numerical variables, including amount, Weight, and height 

```{r}
amount <- bill_amount$amount
weight <- clinical_data$weight
height <- clinical_data$height

# Create a function to plot histogram
plot_histogram <- function(var,var_name){
  min_val <- min(var, na.rm = TRUE)
  max_val <- max(var, na.rm = TRUE)
  title <- paste("Histogram of", var_name, "(Min:", round(min_val, 2), ", Max:", round(max_val, 2), ")")
  
  ggplot() +
  aes(x = var) +
  geom_histogram(boundary = min(var),colour = "black", fill = "lightblue") + 
  labs(x = var_name, title = title)
}

# Create a list of all vectors and plot histogram

numerical_vec_list <- list(amount = amount, weight = weight, height = height)
plots <- mapply(plot_histogram, numerical_vec_list, names(numerical_vec_list), SIMPLIFY = FALSE)

# Display the plots
for (plot in plots) {
  print(plot)
}
```

Explore distribution of few categorical variables (All 1,0 flag variables we will plot as bar and treat as discrete)

**Clinical Data**

```{r}

# Select all variables from clinical_data except few which we have analyzed or do not need to clean
cols_to_exclude <- c("id","date_of_admission","date_of_discharge","weight","height")
bar_cols_clinical <- setdiff(names(clinical_data),cols_to_exclude)
bar_cols_demographics <- c("gender","race","resident_status")

clinical_data[,bar_cols_clinical] <- lapply(clinical_data[bar_cols_clinical], as.factor)
demographics[,bar_cols_demographics] <- lapply(demographics[bar_cols_demographics], as.factor)

plot_list <- list()
plot_bar <- function (var,data){
for (col in var){
  p <- ggplot(data,aes_string(x = col)) + 
    geom_bar(fill = "lightblue", color = "black") + 
    geom_text(stat = "count", aes(label = after_stat(count)),
              position = position_stack(vjust = 0.5)) + 
                labs(title = paste(col),x = "",y = "") + theme_light() 
  plot_list[[col]] <- p
}

for (i in seq(1, length(plot_list),by = 2)) {
  start_index <- i
  end_index <- i + 1
  subset_plots <- plot_list[start_index:end_index]
  grid.arrange(grobs = subset_plots,ncol = 2)
}
}
plot_bar(bar_cols_clinical,clinical_data)

```

**Demographic File**

```{r}
plot_bar(bar_cols_demographics,demographics)
```

#### 2.3.4 Conclusion from data exploration

Based on initial data exploration, it seems that we need to do few processing steps before we merge different files to create one final Analytical data set.

Changes:

* medical_history_hbp: convert yes and no to 1 and 0 respectively  

* admission_date and discharge_date in Clinical_data: convert into appropriate date format to join with billing data  
* gender: in demographics needs to be cleaned by assigning m as Male, and f as Female

* race:Chinese,chinese  and India, Indian to be combined separately

* resident_stats: Singaporean and Singapore citizen to be combined together

* There are few missing data points in columns medical_history_sud , medical_history_tum (we might need to keep in mind while conducting future analysis)


## 3 Data Preprocessing

Here, we will clean the data set which can later be used to final joined data set.

### 3.1 Renaming Columns

To help us keep similar convention across files, we will rename the columns

```{r}
clinical_data <- clinical_data %>% rename_at('id', ~'patient_id')
```

### 3.2 Data formats

Convert dates from clinical and bill tables to date format
```{r}
clinical_data$date_of_admission <- as.Date(clinical_data$date_of_admission, format = "%d/%m/%y")
clinical_data$date_of_discharge <- as.Date(clinical_data$date_of_discharge, format = "%d/%m/%y")
bill_id$date_of_admission <- as.Date(bill_id$date_of_admission, format = "%Y-%m-%d")
demographics$date_of_birth <- as.Date(demographics$date_of_birth, format = "%Y-%m-%d")

```

```{r}
# Lets also check duration of data
min(clinical_data$date_of_admission)
max(clinical_data$date_of_admission)
min(bill_id$date_of_admission)
max(bill_id$date_of_admission)

```


```{r}
nv <- c("cgis_adm","cgis_dis","gaf_lv")
clinical_data[, nv] <- lapply(clinical_data[, nv], function(x) as.numeric(as.character(x)))
```


### 3.3 Join the files to create an Analytical dataset

```{r}
df1 <- left_join(clinical_data,demographics,by = "patient_id")
print(nrow(df1))
```

```{r}
df2 <- left_join(bill_id,bill_amount,by = "bill_id")
print(nrow(df2))
```

Since currently, on same admission, patients could have multiple bills, we will sum the amount for each patient on their date of admission.

```{r}
df3 <- df2 %>% group_by(patient_id,date_of_admission) %>% summarize(total_amount = sum(amount),.groups = 'drop')
```

```{r}
dataset <- left_join(df1,df3,by = c("patient_id","date_of_admission"))
print(nrow(dataset))
```

Number of rows are 3400, which is same as Clinical_date, hence our data is joined correctly. Now we will clean rest of the variables 

### 3.4 Catgories Standardization

```{r}
table(dataset$medical_history_hbp)

dataset <- dataset %>%
  mutate(medical_history_hbp = recode(medical_history_hbp, "No" = "0", "Yes" = "1"))

table(dataset$medical_history_hbp)

table(dataset$gender)

dataset <- dataset %>%
  mutate(gender = recode(gender, "m" = "Male", "f" = "Female"))

table(dataset$gender)

table(dataset$race)

dataset <- dataset %>%
  mutate(race = recode(race, "chinese" = "Chinese", "India" = "Indian"))

table(dataset$race)

table(dataset$resident_status)

dataset <- dataset %>%
  mutate(resident_status = recode(resident_status, "Singaporean" = "Singapore citizen"))

table(dataset$resident_status)

```
### 3.5 Feature Engineering

Here we will create additional features:

* Age at time of admission (Year of admission - Year of Birth)

* Length of Stay in inpatient setting

* Any Symptom of MDD

* Any Treatments

* Number of treatments taken

* First Admission Date
  
```{r}
dataset <- dataset %>%
  mutate(age = year(ymd(date_of_admission)) - year(ymd(date_of_birth)))
plot_histogram(dataset$age, "Age")

```

```{r}
dataset <- dataset %>%
  mutate(los = as.integer(difftime(date_of_discharge, date_of_admission, units = "days")))

plot_histogram(dataset$los, "Length of Stay")

```

```{r}
dataset <- dataset %>%
  rowwise() %>%
  mutate(any_symptom = if_else(any(c_across(starts_with("symptom_")) == 1), 1, 0)) %>%
  ungroup()

table(dataset$any_symptom)
```

```{r}
dataset <- dataset %>%
  rowwise() %>%
  mutate(any_treatment = if_else(any(c_across(starts_with("trt_")) == 1), 1, 0)) %>%
  ungroup()

table(dataset$any_treatment)
table(dataset$trt_adt)

```

```{r}
dataset <- dataset %>%
 mutate(across(starts_with("trt_"), ~ as.numeric(as.character(.)))) %>%
  mutate(number_of_treatments = rowSums(select(., starts_with("trt_")))) %>% 
  mutate(across(starts_with("trt_"), ~ as.factor(.)))
```

```{r}
dataset$number_ge_2 <- ifelse(dataset$number_of_treatments >= 2,1,0)
dataset$number_ge_2 <- as.factor(dataset$number_ge_2)
```

```{r}
dataset <- dataset %>%  group_by(patient_id) %>%
  mutate(min_date_of_admission = min(date_of_admission)) %>% ungroup()

dataset %>% filter(min_date_of_admission == date_of_admission) %>% nrow()
length(unique(dataset$patient_id))
```

Since this matched number of patients, the flag is created correctly

## 4. Cohort Creation

We will first get dataset at patient level by looking only at First Admission 

```{r}

patient_level_data <- dataset %>% filter(min_date_of_admission == date_of_admission) 
patient_level_data$First_admission_flag <- "Total Patients in Baseline"
patient_level_data$any_treatment <- as.factor(patient_level_data$any_treatment)

nrow(patient_level_data)
```
### 4.1 Cohort Steps

Identify Patients with

* To inlcude have adult patients (Age >= 18)

* To exclude patients without treatment (At least 1 Therapy)

* To get moderately ill unresolved MDD patients (cgis at admission >= 4)

```{r}

nrow(patient_level_data)

# Age >= 18

patient_level_data <- patient_level_data %>% filter(age >= 18)
nrow(patient_level_data)

# Any treatmnet taken
patient_level_data <- patient_level_data %>% filter(any_treatment == 1)
nrow(patient_level_data)

# Unresolved MDD
patient_level_data <- patient_level_data %>% filter(cgis_adm >=4)
nrow(patient_level_data)
```

### 4.1 Cohort Demographics and Treatment Statistics

Create table using rtables function


Write a function to create clean tables in R 
```{r}
create_tables <- function(char_vars,num_vars){
lyt <- basic_table(show_colcounts = TRUE) %>%
   split_cols_by("First_admission_flag") %>%
   summarize_row_groups() %>%
   analyze(char_vars,afun = counts_wpcts) %>% 
             analyze(num_vars, function(x, ...) {
    if (is.numeric(x)) {
      in_rows(
        "Mean (sd)" = c(mean(x), sd(x)),
        "Median" = median(x),
        "Min - Max" = range(x),
        .formats = c("xx.xx (xx.xx)", "xx.xx", "xx.xx - xx.xx")
      )
    } else if (is.factor(x) || is.character(x)) {
      in_rows(.list = list_wrap_x(table)(x))
    } else {
      stop("type not supported")
    }
  })
build_table(lyt, patient_level_data)
}
```

Understand demographics

```{r}
create_tables(c("gender","race", "resident_status"),c("age","los","total_amount"))
```


Understand Clinical Characteristics

```{r}
medical_history_columns <- grep("medical_history", names(dataset), value = TRUE)
symptom_columns <- grep("symptom_", names(dataset), value = TRUE)

create_tables(c(medical_history_columns,symptom_columns,"number_ge_2"),
              c("cgis_adm","cgis_dis","gaf_lv","number_of_treatments"))
```

### 4.2 Treatment Combinations at time of First Admission

```{r}
trt_col <- grep("trt_", names(patient_level_data), value = TRUE)
for (col in trt_col){
  print(patient_level_data %>% filter(!!sym(col) == 1) %>% nrow())
}
```

```{r}
patient_level_data1 <- patient_level_data[,trt_col]

patient_level_data1 <- apply(patient_level_data1, 2, as.character)
patient_level_data1 <- apply(patient_level_data1, 2, as.numeric)

trt_df <- patient_level_data %>% group_by(trt_anx,trt_con,trt_adt,trt_ssr,trt_the,trt_oth)
```

```{r}
cross_counts <- crossprod(as.matrix(patient_level_data1))
cross_counts
```

```{r}
diag(cross_counts) <- 0
# Convert cross_counts to a data frame
cross_counts_df <- as.data.frame(cross_counts, stringsAsFactors = FALSE)
cross_counts_df$row <- rownames(cross_counts_df)
cross_counts_df <- cross_counts_df %>%
  gather(col, count, -row) %>%
  mutate(col = factor(col, levels = colnames(cross_counts)))

# Plot heatmap with counts inside
ggplot(cross_counts_df, aes(x = col, y = row, fill = count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = count), size = 4, color = "black") +  # Adjust size here
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal() +
  labs(x = "Treatment", y = "Treatment", title = "Counts Heatmap") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. Final Analysis

Objective:

* Association between Symptoms and treatments


### 5.1 Hypothesis testing

Create a data by removing rows that have NA value
```{r}
patient_level_data_NA_rm <- patient_level_data %>%
  filter(!is.na(medical_history_sud) & !is.na(medical_history_tum))
nrow(patient_level_data_NA_rm)
```

```{r}

patient_level_data2 <- patient_level_data_NA_rm[, c(trt_col, symptom_columns)]

patient_level_data2 <- apply(patient_level_data2, 2, function(x) as.numeric(as.character(x)))

patient_level_data2 <- as.data.frame(patient_level_data2)

trt_symp <- patient_level_data2 %>% 
  group_by(across(all_of(c(trt_col, symptom_columns))))

cross_counts <- crossprod(as.matrix(patient_level_data2))

rows_to_remove <- trt_col
columns_to_remove <- symptom_columns

filtered_cross_counts <- cross_counts[!rownames(cross_counts) %in% rows_to_remove, !colnames(cross_counts) %in% columns_to_remove]
filtered_cross_counts
```
Perform Chi-Square test to check association

```{r}
chisq.test(as.matrix(filtered_cross_counts))
```
The p value is >= 0.05 hence it seems there is no significant association between treatment and symptoms.

### 5.2 Logistic Model

To check association between treatment and medical history by adjusting for factors like demographics and cgis_score, we will see if we get any effect.


```{r}
# Convert Cgis to factor (As we want to treat it as categorical in model)
patient_level_data_NA_rm$cgis_adm <- as.factor(patient_level_data_NA_rm$cgis_adm)
```

```{r}
# Fit logistic regression model
logit_model_anx <- glm(trt_anx ~  cgis_adm + gender + race + resident_status + age + medical_history_dia + medical_history_sud + medical_history_hbp + medical_history_ren + medical_history_tum + medical_history_anx + medical_history_mood, data = patient_level_data_NA_rm, family = binomial)

# Tidy up the model results using broom package
tidy_results_anx <- tidy(logit_model_anx)

# Print adjusted odds ratios and associated statistics
print(tidy_results_anx)
```


```{r}
# Fit logistic regression model
logit_model_con <- glm(trt_con ~  cgis_adm + gender + race + resident_status + age + medical_history_dia + medical_history_sud + medical_history_hbp + medical_history_ren + medical_history_tum + medical_history_anx + medical_history_mood, data = patient_level_data_NA_rm, family = binomial)

# Tidy up the model results using broom package
tidy_results_con <- tidy(logit_model_con)

# Print adjusted odds ratios and associated statistics
print(tidy_results_con)
```

```{r}
# Get the OR and confidence intervals
coef_exp <- exp(coef(logit_model_con))
confint_exp <- exp(confint(logit_model_con))

# Create the dataframe
df_con <- data.frame(
  term = names(coef_exp),
  OR = coef_exp,
  CI_lower = confint_exp[, 1],
  CI_upper = confint_exp[, 2]
)

# Print the dataframe
print(df_con)

# Create the ggplot
ggplot(df_con, aes(x = term, y = OR)) +
  geom_point() +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  scale_y_log10() +  # Log-transform the y-axis for better visualization
  labs(title = "Odds Ratios with 95% Confidence Intervals",
       x = "Variables",
       y = "Odds Ratio (log scale)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


```{r}
# Fit logistic regression model
logit_model_adt <- glm(trt_adt ~  cgis_adm + gender + race + resident_status + age + medical_history_dia + medical_history_sud + medical_history_hbp + medical_history_ren + medical_history_tum + medical_history_anx + medical_history_mood, data = patient_level_data_NA_rm, family = binomial)

# Tidy up the model results using broom package
tidy_results_adt <- tidy(logit_model_adt)

# Print adjusted odds ratios and associated statistics
print(tidy_results_adt)
```

```{r}
# Fit logistic regression model
logit_model_ssr <- glm(trt_ssr ~  cgis_adm + gender + race + resident_status  + age + medical_history_dia + medical_history_sud + medical_history_hbp + medical_history_ren + medical_history_tum + medical_history_anx + medical_history_mood, data = patient_level_data_NA_rm, family = binomial)

# Tidy up the model results using broom package
tidy_results_ssr <- tidy(logit_model_ssr)

# Print adjusted odds ratios and associated statistics
print(tidy_results_ssr)
```




```{r}
# Fit logistic regression model
logit_model_the <- glm(trt_the ~  cgis_adm + gender + race + resident_status  + age + medical_history_dia + medical_history_sud + medical_history_hbp + medical_history_ren + medical_history_tum + medical_history_anx + medical_history_mood, data = patient_level_data_NA_rm, family = binomial)

# Tidy up the model results using broom package
tidy_results_the <- tidy(logit_model_the)

# Print adjusted odds ratios and associated statistics
print(tidy_results_the)
```

```{r}
# Fit logistic regression model
logit_model_oth <- glm(trt_oth ~  cgis_adm + gender + race + resident_status  + age + medical_history_dia + medical_history_sud + medical_history_hbp + medical_history_ren + medical_history_tum + medical_history_anx + medical_history_mood, data = patient_level_data_NA_rm, family = binomial)

# Tidy up the model results using broom package
tidy_results_oth <- tidy(logit_model_oth)

# Print adjusted odds ratios and associated statistics
print(tidy_results_oth)
```


```{r}
# Get the OR and confidence intervals
coef_exp <- exp(coef(logit_model_oth))
confint_exp <- exp(confint(logit_model_oth))

# Create the dataframe
df_oth <- data.frame(
  term = names(coef_exp),
  OR = coef_exp,
  CI_lower = confint_exp[, 1],
  CI_upper = confint_exp[, 2]
)

# Print the dataframe
print(df_oth)

# Create the ggplot
ggplot(df_oth, aes(x = term, y = OR)) +
  geom_point() +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2) +
  scale_y_log10() +  # Log-transform the y-axis for better visualization
  labs(title = "Odds Ratios with 95% Confidence Intervals",
       x = "Variables",
       y = "Odds Ratio (log scale)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### 5.3 Conclusion

* We observe weak relationship between treatments and medical history

* For medical history of anxiety, the odds of prescribing  anticonvulsants increases by factor of 1.28 which was statistically significant
